But seriously: what do algorithms want?

### 2023-01-30 ART_Toscano/Toscano.tex
Toscano article added

\begin{document}
\title{But seriously: what do algorithms want?}
\maketitle

\title{Implying collective intentionalities in algorithmic relays—a distributed cognition approach}
\maketitle

\section*{Abstract}
Describing an algorithm can provide a~formalization of a~specific process. However, different ways of conceptualizing algorithms foreground certain issues while obscuring others. This article attempts to define an algorithm in a~broad sense as a~cultural activity of key importance to make sense of socio-cognitive structures. It also attempts to develop a~sharper account on the interaction between humans and tools, symbols and technologies. Rather than human or machine-centered analyses, I~draw upon sociological and anthropological theories that underline social practices to propose expanding our understanding of an algorithm through the notion of ‘collective intentionalities'. To make this term clear, a~brief historical review is presented, followed by an argumentation on how to incorporate it in an integral perspective. The article responds to recent debates in critical algorithm studies about the significance of the term. It develops a~discussion along the lines of cognitive anthropology and the cognitive sciences, therefore advancing a~definition that is grounded in observed practices as well as in modeled descriptions. The benefit of this approach is that it encourages scholars to explore cognitive structures via archaeologies of technological assemblages, where intentionalities play a~defining role in understanding socio-structured practices and cognitive ecologies.

\section*{Keywords}
algorithm studies, distributed cognition, collective intentionalities, socio-computing infrastructures, cognitive anthropology

@@

On the one hand, it concentrates on processes that happen inside computational machines. This makes the description not only machine-centered, but also introduces a~misunderstanding. After all, articulating a~notion of code in the early days of computing history, pioneer logicians Newell, Simon and Shaw wrote in a~seminal paper that ``the appropriate way to describe a~piece of problem solving behavior is in terms of a~program […]. Computers come into the picture only because they can, by appropriate programming, be induced to execute the same sequences of information processes that humans execute when they are solving problems''
%\label{ref:RNDBmjqIEJ8RS}(Newell, Simon and Shaw, 1958, p.151).
\parencite[][p.151]{newell_elements_1958}. %
 In this sense, as programs, algorithms need not be thought of merely as machine drivers. And as we will see, getting rid of this idiosyncratic constraint would allow us to spot algorithms everywhere, as cultural artifacts 
%\label{ref:RNDhsaVRFFXMo}(Finn, 2017, p.15; Seaver, 2017).
\parencites[][p.15]{finn_what_2017}[][]{seaver_algorithms_2017}.%

@@

If we apply this notion of artifact mediation to our search, we can see that, whether through direct tools or indirect symbols, an algorithm implies an iterative interaction with technology, or in other words, a~practice of recursive intertwining between humans and the technologies they produce. Yet, for Vygotsky, an interaction with symbols or tools is not simply functional, in the sense in which a~subject manipulates an object at will to achieve a~task. Instead, cultural artifacts regulate interactions with one's environment and with oneself. But this is not innocuous: cultural mediation has a~recursive, bidirectional effect; mediated activity simultaneously modifies both the environment and the subject. Cultural mediation influences behaviors, synthetizes experiences from forbearers, and prepares children to acquire specific sets of accumulated memories, as knowledge
%\label{ref:RNDW8eWxTv3J8}(see here also Connerton, 1989).
\parencite[see here also][]{connerton_how_1989}. %
 This co-constitution is what Malafouris, along a~series of cognitive examinations, has termed as metaplasticity 
%\label{ref:RNDg6Z2NrdKR9}(Malafouris, 2010; 2013; 2015).
\parencites**[][]{malafouris_metaplasticity_2010}[][]{malafouris_how_2013}[][]{malafouris_metaplasticity_2015}.%

In the end, cultural mediation—or the ability to think and operate through cultural artifacts—produces historical modes of thinking—i.e. ideologies—and styles of cognition that affect how we learn, think and represent our environment and ourselves. This is what can be described as the notion of distributed cognition
%\label{ref:RNDSg24DwdaBd}(Cole and Engeström, 1993; Gallagher, 2005; 2013).
\parencites*[][]{salomon_cultural-historical_1993}[][]{gallagher_how_2005}[][]{}. %
 Hollan et al. 
%\label{ref:RNDCkpURhADXA}(2000, p.177)
\parencite*[][p.177]{laland_niche_2000} %
 provide a~definition of this process: ``Distributed cognition means more than that cognitive processes are socially distributed across the members of a~group. It is a~broader conception that includes phenomena that emerge in social interactions as well as interactions between people and structure in their environments.'' The notion of distributed cognition has been a~common hypothesis in linguistics and psychology ever since the writings of Vygotsky were published and made available in different translations since the 1970s, but it is by no means the standard model. There are many critics that maintain that, even if aided through different tools, thinking happens basically inside the brain 
%\label{ref:RNDz69682DG3T}(Adams and Aizawa, 2008; 2010; Loh and Kanai, 2016),
\parencites*[][]{adams_bounds_2008}[][]{malafouris_metaplasticity_2010}[][]{loh_how_2016}, %
 or they present situations in which thinking is affected from ``outside'' factors (which Clowes terms as ``the impact thesis'' 
%\label{ref:RND8Ryfr6A8K0}(Clowes, 2019)
\parencite[][]{clowes_screen_2019}%
). We will not deal here with those arguments, since they appear to have a~strong need for an essentialist form of conceptualizing cognition.

@@

\begin{quotation}
[S]ymbols are thus not mere expressions, instrumentalities, or correlates of our biological, psychological, and social existence; they are prerequisites of it. Without men, no culture, certainly; but equally, and more significantly, without culture, no men
%\label{ref:RNDKLhGzqGJ1x}(1973, p.49).
\parencite*[][p.49]{}.%

\end{quotation}
Geertz's formulation found strong empirical evidence among theoretical biologists, for whom the connection between culture and biology implied more than a~simple correlation. As Laland et al.
%\label{ref:RNDOS0n99sbzG}(2000, p.131)
\parencite*[][p.131]{laland_niche_2000} %
 later would claim: ``cultural traits, such as the use of tools, weapons, fire, cooking, symbols, language, and trade, may have played important roles in driving hominid evolution in general and the evolution of the human brain in particular'' 
%\label{ref:RND7shgYMzA5p}(see also Dunbar, 1993; or Aiello and Wheeler, 1995).
\parencites[see also][]{dunbar_coevolution_1993}[or][]{aiello_expensive-tissue_1995}. %
 Nonetheless, when Geertz and other social scientists started confining everything under the domain of ``culture'', throughout the 1980s, the concept became too broad and lost its specific, explanatory power. As Nick Seaver 
%\label{ref:RNDJO9JEsgpIN}(2017, p.4)
\parencite*[][p.4]{seaver_algorithms_2017} %
 writes: ``Its implicit holism and homogenizing, essentialist tendencies seemed politically problematic and ill suited to the conflictual, changing shape of everyday life.'' As a~response, one of the most resourceful attempts in the social sciences to overcome the difficulties brought about by an all-encompassing concept—which was nonetheless useful as a~theoretical compass on a~structural level—was to turn to the study of practicesand symbolic interactions 
%\label{ref:RNDtoIzySTaYH}(Bourdieu, 1972; Certeau, 1984; Blumer, 1986).
\parencites[][]{bourdieu_outline_1972}[][]{certeau_practice_1984}[][]{blumer_symbolic_1986}. %
 Consequently, many sociologists and anthropologists turned from a~vision of a~frame culture as a~unified domain, to the multiplication of sites and cultures, where they could study and map emerging symbolic orders, sometimes coordinated, sometimes conflicting, out of which to make sense of the different layers of social life. This approach left behind the deterministic tone of previous explanations, with their emphasis on rules, models and texts, and began focusing instead on strategies, interests, improvisations and interactional occurrences. Recovering this emphasis, and back to our line of inquiry, Seaver 
%\label{ref:RNDxmbiXf5hFC}(2017, p.5)
\parencite*[][p.5]{seaver_algorithms_2017} %
 provides a~description of an algorithm that is worth mentioning: ``Like other aspects of culture, algorithms are enacted by practices which do not heed a~strong distinction between technical and non-technical concerns, but rather blend them together. In this view, algorithms are not singular technical objects that enter into many different cultural interactions, but are rather unstable objects, culturally enacted by the practices people use to engage with them.''

Seaver highlights the relational aspect of processes, enacted by practices rooted in cultural codes, therefore avoiding both a~subject-centered perspective as well as a~machine-centered view. In that sense, his definition is in line with a~number of interesting theories and methodologies that have emerged in sociology and science and technology studies over the past two decades, for example: actor-networks
%\label{ref:RNDKCObcDcGrt}(Callon, 1986; Latour, 1992; 2005),
\parencites*[][]{law_elements_1986}[][]{bijker_where_1992}[][]{gallagher_how_2005}, %
 sociotechnical ensembles 
%\label{ref:RNDxMte83gBXl}(Bijker, 1999),
\parencite[][]{bijker_bicycles_1999}, %
 object-centered socialities 
%\label{ref:RNDqe3WU3um8T}(Knorr Cetina, 1997),
\parencite[][]{knorr_cetina_sociality_1997}, %
 relational materialities 
%\label{ref:RNDyeyw2YVbeV}(Law, 2004),
\parencite[][]{law_after_2004}, %
 constitutive entanglements 
%\label{ref:RNDJ4mXcd66fz}(Orlikowski, 2007)
\parencite[][]{orlikowski_sociomaterial_2007} %
 or object-oriented ontology 
%\label{ref:RNDSDtw883D9O}(Harman, 2002; Bryant, 2010),
\parencites[][]{harman_tool-being_2002}[][]{bryant_onticologymanifesto_2010}, %
 as well as the approach of cognitive ecology 
%\label{ref:RNDBy4aR50ipy}(Hutchins, 2010)
\parencite[][]{hutchins_cognitive_2010} %
 and material engagement theory 
%\label{ref:RNDdpO41i1cNn}(Malafouris, 2005; 2013)
\parencites*[][]{malafouris_cognitive_2005}[][]{} %
 in the cognitive sciences. These theories challenge and transcend conventional distinctions between objects and subjects, as well as between social abstractions and material iterations. Furthermore, their particular value lies in their insistence on speaking of the social (e.g. culture) and the material (e.g. nature) in the same register, and on not resorting to a~limiting dualism that treats them as separate, even if interacting, phenomena.

@@

\begin{quotation}
Every mental phenomenon is characterized by […] the intentional (or mental) inexistence of an object, and what we might call, though not wholly unambiguously, reference to a~content, direction toward an object (which is not to be understood here as meaning a~thing), or immanent objectivity. Every mental phenomenon includes something as object within itself […] We can, therefore, define mental phenomena by saying that they are those phenomena which contain an object intentionally within themselves.

\end{quotation}
At this point, intentionality was described as a~clear attribute of mental activity, independent of a~real world, but clearly related to it, and decisive to ascribe it meaning. This trait was important because it offered a~form of cognizing reality without relying on the Kantian formulation that attempted to align (individual) sensations and (social) concepts. In other words, it created a~model where things could be cognized beyond a~thick web of structured epistemological pre-conceptions. This is precisely what encouraged Husserl's enthusiasm, as inscribed in his motto ``Back to the things themselves!'' (\textit{Zurück zu den Sachen selbst}!). For as Merleau-Ponty
%\label{ref:RNDs2qs7wId98}(2005, p.xix)
\parencite*[][p.xix]{merleau-ponty_phenomenology_2005} %
 writes: ``What distinguishes intentionality from the Kantian relation to a~possible object is that the unity of the world, before being posited by knowledge in a~specific act of identification, is ‘lived' as ready-made or already there.''

@@

\begin{quotation}
the self appears in experience essentially as a~``me'' with the organization of the community to which it belongs. This organization is, of course, expressed in the particular endowment and particular social situation of the individual […]. He is what he is in so far as he is a~member of this community, and the raw materials out of which this particular individual is born would not be a~self but for his relationship to others in the community of which he is a~part
%\label{ref:RNDT0MXT4rgdz}(Mead, 1972, p.200).
\parencite[][p.200]{mead_mind_1972}.%

\end{quotation}
Following the Vygotsky/Mead hypothesis, there cannot even be a~``direct'' connection between an individual and her experience, because this connection is mediated through language, by which a~``self'' appears as some type of thing. In other words, the emergence of a~``self'' is an effect, or a~functional construction, of a~subject that has learned how to enunciate and use the particle ``I'' under a~given set of socially-sanctioned, grammatical rules. This brings us to a~rather interesting situation on the cognitive side. For if the self is a~social construction, what is to be done with what we call ``the mental''? Is the link between both notions merely a~deficient attribution, or is it a~faulty causal connection? Mead describes the mental as an emergent phenomenon, which involves a~relationship to the character of things:

\begin{quotation}
Those characters are in the things, and while the stimuli call out the response which is in one sense present in the organism, the responses are to things out there. The whole process is not a~mental product and you cannot put it inside of the brain. Mentality is that relationship of the organism to the situation which is mediated by sets of symbols
%\label{ref:RNDaP95DqbXRy}(Mead, 1972, pp.124–125).
\parencite[][pp.124–125]{mead_mind_1972}.%

\end{quotation}
This turns irrelevant the attribution of mentality to the self. On the same grounds, a~causal connection between them can only be inferred as inexistent. Instead, both are equally emergent effects of a~given symbolic mediation. Mead's description of the mental (thatcognitive relationship of an organism to a~situation, mediated by symbols) is the backbone to the definition of an algorithm that was proposed on the previous section. It is also a~touchstone in the tradition of cognitive anthropology that has been associated with the idea of cognitive ecologies
%\label{ref:RNDcmyxI6a8WG}(Douglas, 1986; Lave, 1988; Connerton, 1989; Hutchins, 1995; 2010),
\parencites*[][]{douglas_how_1986}[][]{lave_cognition_1988}[][]{connerton_how_1989}[][]{hutchins_cognition_1995}[][]{menary_defending_2010}, %
 as well as in traditions of cognitive sciences that inquire into models of an embodied, embedded, extended and/or an enactive social mind 
%\label{ref:RNDHOwURfIJuQ}(Clark, 1997; 2003; 2015; Clark and Chalmers, 1998; Gallagher, 2005; 2013; Gallagher and Miyahara, 2012).
\parencites***[][]{clark_being_1997}[][]{}[][]{malafouris_metaplasticity_2015}[][]{clark_extended_1998}[][]{gallagher_how_2005}[][]{malafouris_how_2013}[][]{gallagher_neo-pragmatism_2012}. %
 Gallagher 
%\label{ref:RNDdhebNGSoIx}(2013, p.4),
\parencite*[][p.4]{gallagher_socially_2013}, %
 for instance, describes the mental in this way:

\begin{quotation}
If we think of the mind not as a~repository of propositional attitudes and information, or in terms of internal belief-desire psychology, but as a~dynamic process involved in solving problems and controlling behavior and action—in dialectical, transformative relations with the environment—then we extend our cognitive reach by engaging with tools, technologies, but also with institutions. We create these institutions via our own (shared) mental processes, or we inherit them as products constituted in mental processes already accomplished by others.

\end{quotation}
Indeed, breaking the causal link between the mind and the self allow us to see the dense and emergent network of affordances and enactions that constitute cognitive phenomena. But how do intentionalities come back into the picture? For Brentano, intentionalities were so much as the mark of the mental, i.e. the defining quality of an in-existent, psychological phenomenon. But if the mind is not any more located in an inner, private world, should we just simply do without them? Quite the opposite. As a~matter of fact, intentionalities play a~stronger role within a~distributive cognition approach. But we need to refine the conceptual frame to see how this can be integrated into a~comprehensive explanation.

An intentionality is not a~purpose, nor a~design or an intention to do something, although the notions are closely related. Actions are intentional, for example, not only because there is a~will behind them, but also because they follow a~goal or a~project. If I~am hungry and I~do not have anything to eat at home, I~can go out to a~supermarket to buy groceries in order to cook, or to a~restaurant, or even to a~place where food is distributed if my economic means are limited. These, among others, are available modes of action, connected to material and technical functions, social behaviors and actionable symbolic networks. But we know that there used to be a~time when, if hungry, people could go out hunting or foraging, and the relevant social programs were there to support those activities. Intentionalities are attached then to historical norms, cultural repertoires, social habits, communal values, rituals and many other forms and forces that can be seen to shape an individual's action. For as Brandom
%\label{ref:RNDYwW0p4SJeW}(1994, p.61)
\parencite*[][p.61]{brandom_making_1994} %
 writes: ``only communities, not individuals, can be interpreted as having an original intentionality. [T]he practices that institute the sort of normative status characteristic of intentional states must be social practices.'' In that sense, the social life of an individual consists in a~good deal in determining the appropriateness of her own desires and needs as these are articulated to the available social practices, or cultural programs, through inferential reasoning, practical adjustments and other means.

>>>>

\begin{artengenv}{Javier Toscano}
	{But seriously: what do algorithms want? Implying collective intentionalities in algorithmic relays---a distributed cognition approach}
	{But seriously: what do algorithms want?z\ldots}
	{But seriously: what do algorithms want? Implying collective\\intentionalities in algorithmic relays---a distributed cognition\\approach}
	{Center for Advanced Internet Studies}
	{Describing an algorithm can provide a~formalization of a~specific process. However, different ways of conceptualizing algorithms foreground certain issues while obscuring others. This article attempts to define an algorithm in a~broad sense as a~cultural activity of key importance to make sense of socio-cognitive structures. It also attempts to develop a~sharper account on the interaction between humans and tools, symbols and technologies. Rather than human or machine-centered analyses, I~draw upon sociological and anthropological theories that underline social practices to propose expanding our understanding of an algorithm through the notion of ‘collective intentionalities'. To make this term clear, a~brief historical review is presented, followed by an argumentation on how to incorporate it in an integral perspective. The article responds to recent debates in critical algorithm studies about the significance of the term. It develops a~discussion along the lines of cognitive anthropology and the cognitive sciences, therefore advancing a~definition that is grounded in observed practices as well as in modeled descriptions. The benefit of this approach is that it encourages scholars to explore cognitive structures via archaeologies of technological assemblages, where intentionalities play a~defining role in understanding socio-structured practices and cognitive ecologies.
	}
	{algorithm studies, distributed cognition, collective intentionalities, socio-computing infrastructures, cognitive anthropology.}

%\begin{document}
%\title{But seriously: what do algorithms want?}
%\maketitle
%
%\title{Implying collective intentionalities in algorithmic relays---a distributed cognition approach}
%\maketitle
%
%\section*{Abstract}
%Describing an algorithm can provide a~formalization of a~specific process. However, different ways of conceptualizing algorithms foreground certain issues while obscuring others. This article attempts to define an algorithm in a~broad sense as a~cultural activity of key importance to make sense of socio-cognitive structures. It also attempts to develop a~sharper account on the interaction between humans and tools, symbols and technologies. Rather than human or machine-centered analyses, I~draw upon sociological and anthropological theories that underline social practices to propose expanding our understanding of an algorithm through the notion of ‘collective intentionalities'. To make this term clear, a~brief historical review is presented, followed by an argumentation on how to incorporate it in an integral perspective. The article responds to recent debates in critical algorithm studies about the significance of the term. It develops a~discussion along the lines of cognitive anthropology and the cognitive sciences, therefore advancing a~definition that is grounded in observed practices as well as in modeled descriptions. The benefit of this approach is that it encourages scholars to explore cognitive structures via archaeologies of technological assemblages, where intentionalities play a~defining role in understanding socio-structured practices and cognitive ecologies.
%
%\section*{Keywords}
%algorithm studies, distributed cognition, collective intentionalities, socio-computing infrastructures, cognitive anthropology

@@

On the one hand, it concentrates on processes that happen inside computational machines. This makes the description not only machine-centered, but also introduces a~misunderstanding. After all, articulating a~notion of code in the early days of computing history, pioneer logicians Newell, Simon and Shaw wrote in a~seminal paper that
\myquote{
the appropriate way to describe a~piece of problem solving behavior is in terms of a~program […]. Computers come into the picture only because they can, by appropriate programming, be induced to execute the same sequences of information processes that humans execute when they are solving problems
%\label{ref:RNDBmjqIEJ8RS}(Newell, Simon and Shaw, 1958, p.151).
\parencite[][p.151]{newell_elements_1958}. %
}
 In this sense, as programs, algorithms need not be thought of merely as machine drivers. And as we will see, getting rid of this idiosyncratic constraint would allow us to spot algorithms everywhere, as cultural artifacts 
%\label{ref:RNDhsaVRFFXMo}(Finn, 2017, p.15; Seaver, 2017).
\parencites[][p.15]{finn_what_2017}[][]{seaver_algorithms_2017}.%

@@

If we apply this notion of artifact mediation to our search, we can see that, whether through direct tools or indirect symbols, an algorithm implies an iterative interaction with technology, or in other words, a~practice of recursive intertwining between humans and the technologies they produce. Yet, for Vygotsky, an interaction with symbols or tools is not simply functional, in the sense in which a~subject manipulates an object at will to achieve a~task. Instead, cultural artifacts regulate interactions with one's environment and with oneself. But this is not innocuous: cultural mediation has a~recursive, bidirectional effect; mediated activity simultaneously modifies both the environment and the subject. Cultural mediation influences behaviors, synthetizes experiences from forbearers, and prepares children to acquire specific sets of accumulated memories, as knowledge
%\label{ref:RNDW8eWxTv3J8}(see here also Connerton, 1989).
\parencite[see here also][]{connerton_how_1989}. %
 This co-constitution is what Malafouris, along a~series of cognitive examinations, has termed as metaplasticity 
%\label{ref:RNDg6Z2NrdKR9}(Malafouris, 2010; 2013; 2015).
\parencites[][]{malafouris_metaplasticity_2010}[][]{malafouris_how_2013}[][]{malafouris_metaplasticity_2015}.%

In the end, cultural mediation---or the ability to think and operate through cultural artifacts---produces historical modes of thinking---i.e. ideologies---and styles of cognition that affect how we learn, think and represent our environment and ourselves. This is what can be described as the notion of distributed cognition
%\label{ref:RNDSg24DwdaBd}(Cole and Engeström, 1993; Gallagher, 2005; 2013).
\parencites[][]{salomon_cultural-historical_1993}[][]{gallagher_how_2005}[][]{gallagher_socially_2013}. %
 Hollan et al. 
%\label{ref:RNDCkpURhADXA}(2000, p.177)
\parencite*[][p.177]{laland_niche_2000} %
 provide a~definition of this process: ``Distributed cognition means more than that cognitive processes are socially distributed across the members of a~group. It is a~broader conception that includes phenomena that emerge in social interactions as well as interactions between people and structure in their environments.'' The notion of distributed cognition has been a~common hypothesis in linguistics and psychology ever since the writings of Vygotsky were published and made available in different translations since the 1970s, but it is by no means the standard model. There are many critics that maintain that, even if aided through different tools, thinking happens basically inside the brain 
%\label{ref:RNDz69682DG3T}(Adams and Aizawa, 2008; 2010; Loh and Kanai, 2016),
\parencites[][]{adams_bounds_2008}[][]{menary_defending_2010}[][]{loh_how_2016}, %
 or they present situations in which thinking is affected from ``outside'' factors (which Clowes terms as ``the impact thesis'' 
%\label{ref:RND8Ryfr6A8K0}(Clowes, 2019)
\parencite[][]{clowes_screen_2019}%
). We will not deal here with those arguments, since they appear to have a~strong need for an essentialist form of conceptualizing cognition.

@@

\myquote{
[S]ymbols are thus not mere expressions, instrumentalities, or correlates of our biological, psychological, and social existence; they are prerequisites of it. Without men, no culture, certainly; but equally, and more significantly, without culture, no men
%\label{ref:RNDKLhGzqGJ1x}(1973, p.49).
\parencite*[][p.49]{}.%
}

Geertz's formulation found strong empirical evidence among theoretical biologists, for whom the connection between culture and biology implied more than a~simple correlation. As Laland et al.
%\label{ref:RNDOS0n99sbzG}(2000, p.131)
\parencite*[][p.131]{laland_niche_2000} %
 later would claim: ``cultural traits, such as the use of tools, weapons, fire, cooking, symbols, language, and trade, may have played important roles in driving hominid evolution in general and the evolution of the human brain in particular'' 
%\label{ref:RND7shgYMzA5p}(see also Dunbar, 1993; or Aiello and Wheeler, 1995).
\parencites[see also][]{dunbar_coevolution_1993}[or][]{aiello_expensive-tissue_1995}. %
 Nonetheless, when Geertz and other social scientists started confining everything under the domain of ``culture'', throughout the 1980s, the concept became too broad and lost its specific, explanatory power. As Nick Seaver 
%\label{ref:RNDJO9JEsgpIN}(2017, p.4)
\parencite*[][p.4]{seaver_algorithms_2017} %
 writes: ``Its implicit holism and homogenizing, essentialist tendencies seemed politically problematic and ill suited to the conflictual, changing shape of everyday life.'' As a~response, one of the most resourceful attempts in the social sciences to overcome the difficulties brought about by an all-encompassing concept---which was nonetheless useful as a~theoretical compass on a~structural level---was to turn to the study of practicesand symbolic interactions 
%\label{ref:RNDtoIzySTaYH}(Bourdieu, 1972; Certeau, 1984; Blumer, 1986).
\parencites[][]{bourdieu_outline_1972}[][]{certeau_practice_1984}[][]{blumer_symbolic_1986}. %
 Consequently, many sociologists and anthropologists turned from a~vision of a~frame culture as a~unified domain, to the multiplication of sites and cultures, where they could study and map emerging symbolic orders, sometimes coordinated, sometimes conflicting, out of which to make sense of the different layers of social life. This approach left behind the deterministic tone of previous explanations, with their emphasis on rules, models and texts, and began focusing instead on strategies, interests, improvisations and interactional occurrences. Recovering this emphasis, and back to our line of inquiry, Seaver 
%\label{ref:RNDxmbiXf5hFC}(2017, p.5)
\parencite*[][p.5]{seaver_algorithms_2017} %
 provides a~description of an algorithm that is worth mentioning:
 \myquote{
 Like other aspects of culture, algorithms are enacted by practices which do not heed a~strong distinction between technical and non-technical concerns, but rather blend them together. In this view, algorithms are not singular technical objects that enter into many different cultural interactions, but are rather unstable objects, culturally enacted by the practices people use to engage with them.
 }

Seaver highlights the relational aspect of processes, enacted by practices rooted in cultural codes, therefore avoiding both a~subject-centered perspective as well as a~machine-centered view. In that sense, his definition is in line with a~number of interesting theories and methodologies that have emerged in sociology and science and technology studies over the past two decades, for example: actor-networks
%\label{ref:RNDKCObcDcGrt}(Callon, 1986; Latour, 1992; 2005),
\parencites[][]{law_elements_1986}[][]{bijker_where_1992}[][]{latour_reassembling_2005}, %
 sociotechnical ensembles 
%\label{ref:RNDxMte83gBXl}(Bijker, 1999),
\parencite[][]{bijker_bicycles_1999}, %
 object-centered socialities 
%\label{ref:RNDqe3WU3um8T}(Knorr Cetina, 1997),
\parencite[][]{knorr_cetina_sociality_1997}, %
 relational materialities 
%\label{ref:RNDyeyw2YVbeV}(Law, 2004),
\parencite[][]{law_after_2004}, %
 constitutive entanglements 
%\label{ref:RNDJ4mXcd66fz}(Orlikowski, 2007)
\parencite[][]{orlikowski_sociomaterial_2007} %
 or object-oriented ontology 
%\label{ref:RNDSDtw883D9O}(Harman, 2002; Bryant, 2010),
\parencites[][]{harman_tool-being_2002}[][]{bryant_onticologymanifesto_2010}, %
 as well as the approach of cognitive ecology 
%\label{ref:RNDBy4aR50ipy}(Hutchins, 2010)
\parencite[][]{hutchins_cognitive_2010} %
 and material engagement theory 
%\label{ref:RNDdpO41i1cNn}(Malafouris, 2005; 2013)
\parencites[][]{malafouris_cognitive_2005}[][]{malafouris_how_2013} %
 in the cognitive sciences. These theories challenge and transcend conventional distinctions between objects and subjects, as well as between social abstractions and material iterations. Furthermore, their particular value lies in their insistence on speaking of the social (e.g. culture) and the material (e.g. nature) in the same register, and on not resorting to a~limiting dualism that treats them as separate, even if interacting, phenomena.

@@

\myquote{
Every mental phenomenon is characterized by […] the intentional (or mental) inexistence of an object, and what we might call, though not wholly unambiguously, reference to a~content, direction toward an object (which is not to be understood here as meaning a~thing), or immanent objectivity. Every mental phenomenon includes something as object within itself […] We can, therefore, define mental phenomena by saying that they are those phenomena which contain an object intentionally within themselves.

}
At this point, intentionality was described as a~clear attribute of mental activity, independent of a~real world, but clearly related to it, and decisive to ascribe it meaning. This trait was important because it offered a~form of cognizing reality without relying on the Kantian formulation that attempted to align (individual) sensations and (social) concepts. In other words, it created a~model where things could be cognized beyond a~thick web of structured epistemological pre-conceptions. This is precisely what encouraged Husserl's enthusiasm, as inscribed in his motto ``Back to the things themselves!'' (\textit{Zurück zu den Sachen selbst}!). For as Merleau-Ponty
%\label{ref:RNDs2qs7wId98}(2005, p.xix)
\parencite*[][p.xix]{merleau-ponty_phenomenology_2005} %
 writes:
 \myquote{
 What distinguishes intentionality from the Kantian relation to a~possible object is that the unity of the world, before being posited by knowledge in a~specific act of identification, is ‘lived' as ready-made or already there.
 }

@@

\myquote{
the self appears in experience essentially as a~``me'' with the organization of the community to which it belongs. This organization is, of course, expressed in the particular endowment and particular social situation of the individual […]. He is what he is in so far as he is a~member of this community, and the raw materials out of which this particular individual is born would not be a~self but for his relationship to others in the community of which he is a~part
%\label{ref:RNDT0MXT4rgdz}(Mead, 1972, p.200).
\parencite[][p.200]{mead_mind_1972}.%
}
Following the Vygotsky/Mead hypothesis, there cannot even be a~``direct'' connection between an individual and her experience, because this connection is mediated through language, by which a~``self'' appears as some type of thing. In other words, the emergence of a~``self'' is an effect, or a~functional construction, of a~subject that has learned how to enunciate and use the particle ``I'' under a~given set of socially-sanctioned, grammatical rules. This brings us to a~rather interesting situation on the cognitive side. For if the self is a~social construction, what is to be done with what we call ``the mental''? Is the link between both notions merely a~deficient attribution, or is it a~faulty causal connection? Mead describes the mental as an emergent phenomenon, which involves a~relationship to the character of things:

\myquote{
Those characters are in the things, and while the stimuli call out the response which is in one sense present in the organism, the responses are to things out there. The whole process is not a~mental product and you cannot put it inside of the brain. Mentality is that relationship of the organism to the situation which is mediated by sets of symbols
%\label{ref:RNDaP95DqbXRy}(Mead, 1972, pp.124–125).
\parencite[][pp.124–125]{mead_mind_1972}.%
}
This turns irrelevant the attribution of mentality to the self. On the same grounds, a~causal connection between them can only be inferred as inexistent. Instead, both are equally emergent effects of a~given symbolic mediation. Mead's description of the mental (thatcognitive relationship of an organism to a~situation, mediated by symbols) is the backbone to the definition of an algorithm that was proposed on the previous section. It is also a~touchstone in the tradition of cognitive anthropology that has been associated with the idea of cognitive ecologies
%\label{ref:RNDcmyxI6a8WG}(Douglas, 1986; Lave, 1988; Connerton, 1989; Hutchins, 1995; 2010),
\parencites[][]{douglas_how_1986}[][]{lave_cognition_1988}[][]{connerton_how_1989}[][]{hutchins_cognition_1995}[][]{hutchins_cognitive_2010}, %
 as well as in traditions of cognitive sciences that inquire into models of an embodied, embedded, extended and/or an enactive social mind 
%\label{ref:RNDHOwURfIJuQ}(Clark, 1997; 2003; 2015; Clark and Chalmers, 1998; Gallagher, 2005; 2013; Gallagher and Miyahara, 2012).
\parencites[][]{clark_being_1997}[][]{clark_natural-born_2003}[][]{clark_what_2015}[][]{clark_extended_1998}[][]{gallagher_how_2005}[][]{gallagher_socially_2013}[][]{gallagher_neo-pragmatism_2012}. %
 Gallagher 
%\label{ref:RNDdhebNGSoIx}(2013, p.4),
\parencite*[][p.4]{gallagher_socially_2013}, %
 for instance, describes the mental in this way:

\myquote{
If we think of the mind not as a~repository of propositional attitudes and information, or in terms of internal belief-desire psychology, but as a~dynamic process involved in solving problems and controlling behavior and action---in dialectical, transformative relations with the environment---then we extend our cognitive reach by engaging with tools, technologies, but also with institutions. We create these institutions via our own (shared) mental processes, or we inherit them as products constituted in mental processes already accomplished by others.

}
Indeed, breaking the causal link between the mind and the self allow us to see the dense and emergent network of affordances and enactions that constitute cognitive phenomena. But how do intentionalities come back into the picture? For Brentano, intentionalities were so much as the mark of the mental, i.e. the defining quality of an in-existent, psychological phenomenon. But if the mind is not any more located in an inner, private world, should we just simply do without them? Quite the opposite. As a~matter of fact, intentionalities play a~stronger role within a~distributive cognition approach. But we need to refine the conceptual frame to see how this can be integrated into a~comprehensive explanation.

An intentionality is not a~purpose, nor a~design or an intention to do something, although the notions are closely related. Actions are intentional, for example, not only because there is a~will behind them, but also because they follow a~goal or a~project. If I~am hungry and I~do not have anything to eat at home, I~can go out to a~supermarket to buy groceries in order to cook, or to a~restaurant, or even to a~place where food is distributed if my economic means are limited. These, among others, are available modes of action, connected to material and technical functions, social behaviors and actionable symbolic networks. But we know that there used to be a~time when, if hungry, people could go out hunting or foraging, and the relevant social programs were there to support those activities. Intentionalities are attached then to historical norms, cultural repertoires, social habits, communal values, rituals and many other forms and forces that can be seen to shape an individual's action. For as Brandom
%\label{ref:RNDYwW0p4SJeW}(1994, p.61)
\parencite*[][p.61]{brandom_making_1994} %
 writes: 
 \myquote{
 only communities, not individuals, can be interpreted as having an original intentionality. [T]he practices that institute the sort of normative status characteristic of intentional states must be social practices.
 }
 In that sense, the social life of an individual consists in a~good deal in determining the appropriateness of her own desires and needs as these are articulated to the available social practices, or cultural programs, through inferential reasoning, practical adjustments and other means.

Now, even if at a~first look this explanation seems to restrain an individual's agency, by making her guide a~certain ``intended'' action through a~given catalogue of socially sanctioned paths, the picture that this model enables is in fact richer and more complex. In a~few words, a~strict functionalism does not apply
%\label{ref:RND4RkvJYvKWK}(Elster, 1983; Douglas, 1986, p.32ff).
\parencites[][]{elster_explaining_1983}[][p.32ff]{douglas_how_1986}. %
 As a~matter of fact, a~model like Malafouris' material engagement theory actually sustains that the distinctive forms of human agency emerge precisely in the practical space afforded by the interactions 
%\label{ref:RNDTsgW80JhWZ}(Malafouris, 2008; 2015).
\parencites[][]{malafouris_at_2008}[][]{malafouris_metaplasticity_2015}. %
 After all, an individual never ``acts'' in a~void either. And as Cooren et al 
%\label{ref:RND8KlhTAyYvS}(2006, p.11)
\parencite*[][p.11]{cooren_communication_2006} %
 write:
 \myquote{
 Agency is not a~‘capacity to act' to be defined a~priori. On the contrary, it is ‘the capacity to act' that is discovered when studying how worlds become constructed in a~certain way.
 }
 In that sense, intentionalities are sustained in social practices without losing their capacity for an individual's adaptation, expression and further innovation. And as such, they can be acknowledged as \textit{collective intentionalities}, fundamental pieces that connect an individual to a~larger collective, without necessarily turning them into a~deterministic setup. Collective intentionalities are in that sense something as action-able paths, through which an individual orients and articulates her actions with the resources and experiences of a~cultural community, i.e. a~community of practice.
